---
title: "Weedy Rice ICP-MS"
author: "Fengling Hu"
output: pdf_document
classoption: landscape
---
load data in
```{r, include = FALSE}
dyn.load('/Library/Java/JavaVirtualMachines/jdk1.8.0_144.jdk/Contents/Home/jre/lib/server/libjvm.dylib')
require(xlsx); require(tidyverse); require(reshape); require(mosaic); require(matrixStats)
setwd("/Users/hufengling/OneDrive - Amherst College/Documents/Amherst/Jeong Research/Weedy Rice")

#read files (rr20parent was removed from Excel Combined files because it is probably contaminated)
ms1 <- read.xlsx("041217 Combined.xlsx", 1) %>% dplyr::rename(p = p31cps, mn = mn55ppb, fe = fe57ppb, zn = zn66ppb, as = as75ppb, cd = cd111ppb, pb = pb208ppb)
ms2 <- read.xlsx("050517 Combined.xlsx", 1) %>% dplyr::rename(p = p31cps, mn = mn55ppb, fe = fe57ppb, zn = zn66ppb, as = as75ppb, cd = cd111ppb, pb = pb208ppb)
ms3 <- read.xlsx("062317 Combined.xlsx", 1) %>% dplyr::rename(p = p31cps, mn = mn55ppb, fe = fe57ppb, zn = zn66ppb, as = as75ppb, cd = cd111ppb, pb = pb208ppb)
ms4 <- read.xlsx("092717 Combined.xlsx", 1) %>% dplyr::rename(p = p31cps, mn = mn55ppb, fe = fe57ppb, zn = zn66ppb, as = as75ppb, cd = cd111ppb, pb = pb208ppb)
ms5 <- read.xlsx("101717 Combined.xlsx", 1) %>% dplyr::rename(p = p31cps, mn = mn55ppb, fe = fe57ppb, zn = zn66ppb, as = as75ppb, cd = cd111ppb, pb = pb208ppb)

icpmsRaw <- rbind(ms1, ms2, ms3, ms4, ms5) #join all data together

counts <- summary(icpmsRaw[icpmsRaw$Blank == "N",]$ID)
counts <- counts[counts != 0]
counts <- counts %>% as.data.frame()
counts <- counts %>% cbind(ID = rownames(counts))
colnames(counts) <- c("numTrials", "ID")
```

normalize by subtracting trial-specific blanks
```{r, include = FALSE}
#Use graphs to check whether to normalize ms4 for MilliQ water
msblank <- filter(icpmsRaw, Blank == "Y")
msblank <- msblank[-c(5,9,12),] #remove weird points

xyplot(p~Trial, data = msblank)
xyplot(mn~Trial, data = msblank)
xyplot(fe~Trial, data = msblank)
xyplot(zn~Trial, data = msblank)
xyplot(as~Trial, data = msblank)
xyplot(cd~Trial, data = msblank)
xyplot(pb~Trial, data = msblank)

xyplot(p~mq, data = msblank[msblank$Trial == 4,], type = c("p","r"))
xyplot(mn~mq, data = msblank[msblank$Trial == 4,], type = c("p","r"))
xyplot(fe~mq, data = msblank[msblank$Trial == 4,], type = c("p","r"))
xyplot(zn~mq, data = msblank[msblank$Trial == 4,], type = c("p","r"))
xyplot(as~mq, data = msblank[msblank$Trial == 4,], type = c("p","r"))
xyplot(cd~mq, data = msblank[msblank$Trial == 4,], type = c("p","r"))
xyplot(pb~mq, data = msblank[msblank$Trial == 4,], type = c("p","r"))

# ggplot(msblank[msblank$Trial == 4,]) + geom_point(aes(mq,p, color = Trial)) + expand_limits(y=0)
# ggplot(msblank[msblank$Trial == 4,]) + geom_point(aes(mq,mn, color = Trial)) + expand_limits(y=0)
# ggplot(msblank[msblank$Trial == 4,]) + geom_point(aes(mq,fe, color = Trial)) + expand_limits(y=0)
# ggplot(msblank[msblank$Trial == 4,]) + geom_point(aes(mq,zn, color = Trial)) + expand_limits(y=0)
# ggplot(msblank[msblank$Trial == 4,]) + geom_point(aes(mq,as, color = Trial)) + expand_limits(y=0)
# ggplot(msblank[msblank$Trial == 4,]) + geom_point(aes(mq,cd, color = Trial)) + expand_limits(y=0)
# ggplot(msblank[msblank$Trial == 4,]) + geom_point(aes(mq,pb, color = Trial)) + expand_limits(y=0)

#Looks different sometimes. Let's normalize
dimq <- vector("list", 7)
dimq[[1]] <- (lm(p~mq, data = msblank[msblank$Trial == 4,]) %>% summary())[[4]]
dimq[[2]] <- (lm(mn~mq, data = msblank[msblank$Trial == 4,]) %>% summary())[[4]]
dimq[[3]] <- (lm(fe~mq, data = msblank[msblank$Trial == 4,]) %>% summary())[[4]]
dimq[[4]] <- (lm(zn~mq, data = msblank[msblank$Trial == 4,]) %>% summary())[[4]]
dimq[[5]] <- (lm(as~mq, data = msblank[msblank$Trial == 4,]) %>% summary())[[4]]
dimq[[6]] <- (lm(cd~mq, data = msblank[msblank$Trial == 4,]) %>% summary())[[4]]
dimq[[7]] <- (lm(pb~mq, data = msblank[msblank$Trial == 4,]) %>% summary())[[4]]

#calculate estimates for metal concentrations in DI and MQ water based on linear regression
dimq0 <- sapply(dimq, function(x) x[1,1]) %>% t()
dimq100 <- sapply(dimq, function(x) x[1,1] + x[2,1]*100) %>% t()

avgblank <- sapply(c(1:3,5), function(x) colMedians(as.matrix(msblank[msblank$Trial == x, 3:9]))) %>% t()

#Normalize on blanks! ms#B represent datasets normalized for blanks
ms1B <- ms1
ms1B[,3:9] <- apply(ms1[,3:9], 1, function(x) x - avgblank[1,]) %>% t()
ms2B <- ms2
ms2B[,3:9] <- apply(ms2[,3:9], 1, function(x) x - avgblank[2,]) %>% t()
ms3B <- ms3
ms3B[,3:9] <- apply(ms3[,3:9], 1, function(x) x - avgblank[3,]) %>% t()
ms4B <- ms4

for(i in 1:nrow(ms4)){
  x <- ms4[i,]
  if(x[19]==100)
    x[3:9] <- x[3:9]-dimq100
  else if(x[19]==0)
    x[3:9] <- x[3:9]-dimq0
  ms4B[i,] <- x
}

ms5B <- ms5
ms5B[,3:9] <- apply(ms5[,3:9], 1, function(x) x - avgblank[4,]) %>% t()

rm(avgblank, dimq, dimq0, dimq100, msblank, icpmsRaw, x, i)
```

normalize by dividing by mass and multiplying by total dilution volume
```{r, include = FALSE}
#Normalize on masses and dilutions! ms#N represent completely normalized datasets
ms1N <- ms1B
ms2N <- ms2B
ms3N <- ms3B
ms4N <- ms4B
ms5N <- ms5B

#convert to units ppb/mg
ms1N[,3:9] <- sapply(3:9, function(x) ms1B[,x]*((ms1B[,16] + ms1B[,18])*ms1B[,17]/ms1B[,1]))
ms2N[,3:9] <- sapply(3:9, function(x) ms2B[,x]*((ms2B[,16] + ms2B[,18])*ms2B[,17]/ms2B[,1]))
ms3N[,3:9] <- sapply(3:9, function(x) ms3B[,x]*((ms3B[,16] + ms3B[,18])*ms3B[,17]/ms3B[,1]))
ms4N[,3:9] <- sapply(3:9, function(x) ms4B[,x]*((ms4B[,16] + ms4B[,18])*ms4B[,17]/ms4B[,1]))
ms5N[,3:9] <- sapply(3:9, function(x) ms5B[,x]*((ms5B[,16] + ms5B[,18])*ms5B[,17]/ms5B[,1]))

rm(ms1B, ms2B, ms3B, ms4B, ms5B , ms1, ms2, ms3, ms4, ms5)
```

normalize by converting to Z score, scaled to Z-score by common samples run in all trials
```{r, include = FALSE}
#join data
icpmsNorm <- rbind(ms1N, ms2N, ms3N, ms4N, ms5N)
icpmsNorm <- left_join(icpmsNorm, counts, by = "ID")

colnames(icpmsNorm) <- c("mass","id","p","mn","fe","zn","as","cd","pb","trial","seeds","filtered","husks","parent","blank","acid","dil","peroxide","mq","numTrials")
icpmsNorm$trial <- as.factor(icpmsNorm$trial)

commonIDs <- icpmsNorm %>% group_by(id) %>% summarize(trialTotal = sum(trial %>% as.numeric()))
commonIDs <- commonIDs[commonIDs$trialTotal == 15, 1]
commonIDs <- cbind(commonIDs, common = rep("Y", nrow(commonIDs)))

icpmsNorm <- left_join(icpmsNorm, commonIDs, by = "id")

#calculate mean and variance of common samples run in all trials
trialSummary <- (icpmsNorm[icpmsNorm$common == "Y", ] %>% group_by(trial) %>% summarize(pTrialMean = mean(p),mnTrialMean = mean(mn),feTrialMean = mean(fe),znTrialMean = mean(zn),asTrialMean = mean(as),cdTrialMean = mean(cd),pbTrialMean = mean(pb),pTrialSD = sd(p),mnTrialSD = sd(mn),feTrialSD = sd(fe),znTrialSD = sd(zn),asTrialSD = sd(as),cdTrialSD = sd(cd),pbTrialSD = sd(pb)))[1:5,2:15]

#scale to z-score
scaleCommon <- function(df, n){
  temp <- matrix(nrow = nrow(df), ncol = 7)
  for(i in 3:9){
    temp[,i-2] <- (df[,i]-rep(trialSummary[n,i-2] %>% as.numeric(),nrow(df)))/(rep(trialSummary[n,i+5] %>% as.numeric(),nrow(df)))
  }
  temp
}

ms1N <- ms1N %>% cbind(scaleCommon(ms1N,1))
ms2N <- ms2N %>% cbind(scaleCommon(ms2N,2))
ms3N <- ms3N %>% cbind(scaleCommon(ms3N,3))
ms4N <- ms4N %>% cbind(scaleCommon(ms4N,4))
ms5N <- ms5N %>% cbind(scaleCommon(ms5N,5))

#scale by mean and sd of each trial
# ams1 <- lapply(3:9, function(x) ms1N[,x] %>% scale())
# ams2 <- sapply(3:9, function(x) ms2N[,x] %>% scale())
# ams3 <- sapply(3:9, function(x) ms3N[,x] %>% scale())
# ams4 <- sapply(3:9, function(x) ms4N[,x] %>% scale())
```

reformat datasets
```{r, include = FALSE}
#join data
icpms <- rbind(ms1N, ms2N, ms3N, ms4N, ms5N)
icpms <- left_join(icpms, counts, by = "ID")

colnames(icpms) <- c("mass","id","p","mn","fe","zn","as","cd","pb","trial","seeds","filtered","husks","parent","blank","acid","dil","peroxide","mq","pZ","mnZ","feZ","znZ","asZ","cdZ","pbZ", "numTrials")
icpms$trial <- as.factor(icpms$trial)
icpms$numTrials <- as.factor(icpms$numTrials)

icpms <- icpms[icpms$blank == "N",]
icpms$mq <- as.factor(icpms$mq)
icpms <- icpms[order(icpms$id),]

rm(ams1,ams2,ams3,ams4,ams5,ms1B,ms2B,ms3B,ms4B,ms5B,ms1,ms2,ms3,ms4,ms5,ms1N,ms2N,ms3N,ms4N,ms5N,trialSummary, scaleCommon)
```

Looking for Batch Effects

```{r, echo = FALSE, fig.height = 7, fig.width = 10}
#looking for obvious batch effect
ggplot(icpms, aes(trial, p, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Phosphorous normalized ppb by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, mn, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Manganese normalized ppb by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, fe, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Iron normalized ppb by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, zn, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Zinc normalized ppb by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, as, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Arsenic normalized ppb by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, cd, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Cadmium normalized ppb by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, pb, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Lead normalized ppb by Trial by MilliQ/DI")

#trying to control for batch effect using scaling (conversion to z-score)
ggplot(icpms, aes(trial, pZ, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0)+ ggtitle("Phosphorous Z-scores by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, mnZ, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Manganese Z-scores by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, feZ, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Iron Z-scores by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, znZ, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Zinc Z-scores by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, asZ, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Arsenic Z-scores by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, cdZ, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Cadmium Z-scores by Trial by MilliQ/DI")
ggplot(icpms, aes(trial, pbZ, color = mq)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = .2, height = 0) + ggtitle("Lead Z-scores by Trial by MilliQ/DI")
```

Looking at metal content (ppb/mg) by trial per ID

```{r, echo = FALSE, fig.width = 10, fig.height = 7}
#visualizing individuals for all elements
ggplot(icpms, aes(id, p)) + geom_boxplot() + geom_point(aes(id, p, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Phosphorous (ppb) by ID")
ggplot(icpms, aes(id, mn)) + geom_boxplot() + geom_point(aes(id, mn, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Manganese (ppb) by ID")
ggplot(icpms, aes(id, fe)) + geom_boxplot() + geom_point(aes(id, fe, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Iron (ppb) by ID")
ggplot(icpms, aes(id, zn)) + geom_boxplot() + geom_point(aes(id, zn, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Zinc (ppb) by ID")
ggplot(icpms, aes(id, as)) + geom_boxplot() + geom_point(aes(id, as, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Arsenic (ppb) by ID")
ggplot(icpms, aes(id, cd)) + geom_boxplot() + geom_point(aes(id, cd, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Cadmium (ppb) by ID")
ggplot(icpms, aes(id, pb)) + geom_boxplot() + geom_point(aes(id, pb, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Lead (ppb) by ID")
```

Looking at metal content (Z-score) by trial per ID

```{r, echo = FALSE, fig.width = 10, fig.height = 7}
#visualizing individuals for all elements
ggplot(icpms, aes(id, pZ)) + geom_boxplot() + geom_point(aes(id, pZ, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Phosphorous (Z) by ID")
ggplot(icpms, aes(id, mnZ)) + geom_boxplot() + geom_point(aes(id, mnZ, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Manganese (Z) by ID")
ggplot(icpms, aes(id, feZ)) + geom_boxplot() + geom_point(aes(id, feZ, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Iron (Z) by ID")
ggplot(icpms, aes(id, znZ)) + geom_boxplot() + geom_point(aes(id, znZ, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Zinc (Z) by ID")
ggplot(icpms, aes(id, asZ)) + geom_boxplot() + geom_point(aes(id, asZ, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Arsenic (Z) by ID")
ggplot(icpms, aes(id, cdZ)) + geom_boxplot() + geom_point(aes(id, cdZ, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Cadmium (Z) by ID")
ggplot(icpms, aes(id, pbZ)) + geom_boxplot() + geom_point(aes(id, pbZ, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Lead (Z) by ID")

rm(bms1, bms2, bms3, bms4, commonIDs, counts, icpmsNorm, icpmsRaw)
```

Looking at summary measures multiple trials and uniqueness

```{r, echo = FALSE, fig.height = 7, fig.width = 10}
#collecting all trials as averages
medianZs <- icpms %>% group_by(id) %>% summarize(medianpZ = median(pZ), medianmnZ = median(mnZ), medianfeZ = median(feZ), medianznZ = median(znZ), medianasZ = median(asZ), mediancdZ = median(cdZ), medianpbZ = median(pbZ)) %>% mutate(absSum = abs(medianpZ) + abs(medianmnZ) + abs(medianfeZ) + abs(medianznZ) + abs(medianasZ) + abs(mediancdZ) + abs(medianpbZ), sqSum = medianpZ^2 + medianmnZ^2 + medianfeZ^2 + medianznZ^2 + medianasZ^2 + mediancdZ^2 + medianpbZ^2) %>% as.data.frame()

#reshaping dataset
medianZmelt <- melt(medianZs, id = "id", variable_name = "measures")
medianZmeltvars <- medianZmelt[medianZmelt$measures != "absSum",]
medianZmeltvars <- medianZmeltvars[medianZmeltvars$measures != "sqSum",]

#average for all metals across each id
ggplot(medianZmeltvars) + geom_point(aes(id, value, color = measures)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Median Z-scores (across all trials) of all metals by ID\n (larger value represents more overall deviance from the norm")

#Sum of absolute values of mean Z scores
ggplot(medianZmelt[medianZmelt$measures == "absSum",]) + geom_point(aes(id, value)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Sum of absolute values of median Z-scores for all elements by ID\n (larger value represents more overall deviance from the norm)")

ggplot(medianZmelt[medianZmelt$measures == "sqSum",]) + geom_point(aes(id, value)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Sum of squares of median Z-scores for all elements by ID\n (larger value represents more overall deviance from the norm)")

rm(medianZs, medianZmelt, medianZmeltvars, commonIDs, counts)
```

building dataframe to compare husk vs no husk

```{r}
#sort by accession
icpmsAcc <- icpms[icpms$trial != 5,] %>% group_by(id) %>% summarize(medianp = median(p), medianmn = median(mn), medianfe = median(fe), medianzn = median(zn), medianas = median(as), mediancd = median(cd), medianpb = median(pb), medianpZ = median(pZ), medianmnZ = median(mnZ), medianfeZ = median(feZ), medianznZ = median(znZ), medianasZ = median(asZ), mediancdZ = median(cdZ), medianpbZ = median(pbZ)) %>% as.data.frame()

icpms4 <- icpms[icpms$trial == 4,]
icpms5 <- icpms[icpms$trial == 5,]

#join median for trial 1-4, values for trial 4, and values for trial 5 together
icpmshusk <- left_join(icpms4, icpms5, by = c("id", "parent"))[,c(2:9, 20:26, 29:35, 45:51)] %>% left_join(icpmsAcc, ., by = "id")

#check how good trial 4 value and median line up
huskmelt <- melt(icpmshusk)

bytype <- huskmelt[1:73,1] %>% as.data.frame()

for(i in 1:6){
  bytype <- bytype %>% cbind(huskmelt[(i*511-510):(i*511),3])
}

colnames(bytype) <- c("id", "median", "medianZ", "trial4", "trial4Z", "trial5", "trial5Z")

#comparing median to trial4
xyplot(medianZ^(1/3)~trial4Z^(1/3), data = bytype, type = c("p","r"))
bytype <- bytype %>% mutate(medianZt = medianZ^(1/3), trial4Zt = trial4Z^(1/3))
lm(medianZt~trial4Z^(1/3), data = bytype) %>% summary()
```

plot trial5 data to trial4/median data (UNFINISHED)
```{r, echo = FALSE, fig.width = 10, fig.height = 7}
#visualizing individuals for all elements
ggplot(icpmshusk) + geom_point(aes(id, medianp, color = "blue")) +geom_point(aes(id, p.y, color = "red")) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Phosphorous (ppb) by ID")
ggplot(icpms, aes(id, mn)) + geom_boxplot() + geom_point(aes(id, mn, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Manganese (ppb) by ID")
ggplot(icpms, aes(id, fe)) + geom_boxplot() + geom_point(aes(id, fe, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Iron (ppb) by ID")
ggplot(icpms, aes(id, zn)) + geom_boxplot() + geom_point(aes(id, zn, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Zinc (ppb) by ID")
ggplot(icpms, aes(id, as)) + geom_boxplot() + geom_point(aes(id, as, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Arsenic (ppb) by ID")
ggplot(icpms, aes(id, cd)) + geom_boxplot() + geom_point(aes(id, cd, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Cadmium (ppb) by ID")
ggplot(icpms, aes(id, pb)) + geom_boxplot() + geom_point(aes(id, pb, color = trial)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Normalized Lead (ppb) by ID")
```

